---
title: "Data Preparation: Image Differencing Pipeline"
subtitle: "Internal reference — not presented in lab"
author: "Elisabetta Pietrostefani, University of Liverpool"
date: "2026-05-19"
format:
  html:
    code-fold: false
    toc: true
execute:
  eval: false   # set to true when running locally with full MAXAR data
---

::: {.callout-warning}
## Internal document
This file documents the full data preparation pipeline. It is **not part of the student-facing lab**. It requires the full MAXAR `.tif` files (~several GB) to run. Set `eval: true` in the YAML when running locally.

Source: adapted from `GIS_Final_Coursework.Rmd` (Moatti, 2021).
Original repository: [github.com/SaraMoatti/Beirut_explosion](https://github.com/SaraMoatti/Beirut_explosion)
:::

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(terra)
library(sf)
library(tidyverse)
library(here)
```

## Step 1: Load operational zones

```{r}
#| label: load-zones

# Beirut operational zones — defines the study area boundary
zones <- st_read(
  here("data", "geo",
       "beirut_port_explosions_operational_zones_139",
       "beirut_port_explosions_operational_zones_139.shp")
)

#|>
#  st_transform(22770)   # Deir el Zor (EPSG:22770) — local Lebanese projection

plot(st_geometry(zones), main = "Beirut operational zones")
```

## Step 2: Load pre- and post-event MAXAR tiles

```{r}
#| label: load-rasters

# Pre-event WorldView tile (July 2020)
# Source: data/maxar_raw/Before/10300500A5F95600.tif
r_before <- terra::rast(
  here("data", "maxmar", "10300500A5F95600.tif")
)

# Post-event WorldView tile (August 5, 2020)
# Source: data/maxar_raw/After/104001005BD06800.tif
r_after <- terra::rast(
  here("data", "maxmar", "104001005BD06800.tif")
)

# Inspect both
r_before
r_after
```

## Step 3: Align extents

```{r}
#| label: align-extents

# Check extents — they must match before differencing
ext(r_before)
ext(r_after)

# Resample post to match pre if extents differ
# method = "near" (nearest neighbour) preserves pixel values without interpolation
r_after_matched <- terra::resample(r_after, r_before, method = "near")
```

## Step 4: Crop and mask to study area

```{r}
#| label: crop-mask

zones_vect <- terra::vect(zones)

r_before_crop <- r_before |>
  terra::crop(zones_vect) |>
  terra::mask(zones_vect)

r_after_crop <- r_after_matched |>
  terra::crop(zones_vect) |>
  terra::mask(zones_vect)

# Check memory — these are large objects
cat("Before crop size: ", terra::ncell(r_before_crop) * terra::nlyr(r_before_crop), "cells\n")
cat("After crop size:  ", terra::ncell(r_after_crop)  * terra::nlyr(r_after_crop),  "cells\n")
```

## Step 5: Log-ratio differencing (primary output)

```{r}
#| label: log-diff

# log10(after / before) — more robust to illumination differences between dates
# positive = brighter after (rubble, dust, exposed concrete)
# negative = darker after (collapsed structure, shadow)
# near zero = no change

stack_after_before_log <- log10(r_after_crop / r_before_crop)

terra::writeRaster(
  stack_after_before_log,
  filename  = here("data", "maxmar_processed", "stack_after_before_log.tif"),
  overwrite = TRUE
)
```
```{r}
terra::nlyr(stack_after_before_log)  
terra::global(stack_after_before_log, "range", na.rm = TRUE)
terra::global(stack_after_before_log, fun = function(x) sum(!is.finite(x)))
summary(terra::values(stack_after_before_log[[1]]))
```

stack_plot <- stack_after_before_log
stack_plot[!is.finite(stack_plot)] <- NA
stack_plot <- terra::trim(stack_plot)
stack_plot <- terra::clamp(stack_plot, lower = -1, upper = 1, values = TRUE)

plotRGB(stack_plot, r = 1, g = 2, b = 3,
        stretch = "hist", axes = TRUE,
        main = "Log-ratio (after / before)")
        
## Step 6: Visualise the log-ratio difference image

```{r}
#| label: plot-log-diff

# Sanitise for display: drop non-finite (Inf/NaN from divisions by zero),
# clamp tails to ±1 so outliers don't compress the visible range,
# and trim the NA border so the data fills the plot extent.
stack_plot <- stack_after_before_log
stack_plot[!is.finite(stack_plot)] <- NA
stack_plot <- terra::clamp(stack_plot, lower = -1, upper = 1, values = TRUE)
stack_plot <- terra::trim(stack_plot)

# Save as PNG for use in the lab document
# (avoids loading the large raster in the rendered lab)
png(here("data", "stack_after_before_log.png"),
    width = 1200, height = 1000, res = 150)
plotRGB(stack_plot, r = 1, g = 2, b = 3,
        stretch = "hist",
        main    = "Log-ratio difference: Beirut blast (Aug 2020)")
dev.off()

# Interactive preview
plotRGB(stack_plot,
        r = 1, g = 2, b = 3,
        axes    = TRUE,
        stretch = "hist",
        main    = "Log-ratio difference image (after / before)")
```

::: {.callout-note title="Reading the log-ratio RGB image"}

Each pixel in `stack_after_before_log` holds three values — `log10(after / before)` computed independently for bands 1, 2, and 3 of the Maxar multispectral product (approximately Red, Green, Blue). These describe how each colour channel changed between the pre-blast and post-blast acquisitions.

**Sign convention:**

- **Positive** (after > before) — the surface got *brighter* in that band. Typical causes: exposed concrete, rubble, dust deposits, debris fields, newly bare ground where a roof or vegetation used to be.
- **Negative** (after < before) — the surface got *darker*. Typical causes: collapsed structures (cavities, shadow), water intrusion, scorching, loss of bright roofing material.
- **Near zero** — no meaningful change. Stable surfaces, undamaged roofs, roads outside the blast footprint.

**What `plotRGB` does with the values.** Band 1's log-ratio maps to the red channel of the display, band 2 to green, band 3 to blue. `stretch = "hist"` then rescales each band independently across 0–255 by histogram equalisation. The `clamp(-1, 1)` step beforehand pulls extreme outliers (anything beyond a 10× brightness change) to the limits so they don't compress all the meaningful variation into a narrow grey band.

**How to read the colours.** Because real physical change tends to move all three bands together — rubble brightens R, G, and B; collapse darkens all three — the most damaged areas typically show as near-greyscale: bright white/grey for debris and rubble, dark grey/black for collapse and shadow. Strongly *coloured* pixels (saturated red, green, or blue) indicate change in one band but not the others, which is usually spectrally selective change (e.g. vegetation loss) or inter-date noise rather than blast damage.

**Spatial pattern.** The blast crater at the port (north end of the AOI, near the grain silos) shows the most extreme signature. A roughly circular pattern of bright pixels marks the debris and damage zone, fading to near-zero in undamaged areas further out. The operational zones masked in Step 4 define the outer boundary — everything beyond renders as the plot background.

**Caveat.** `stretch = "hist"` equalises *within the displayed extent*, so the colours are relative to this crop. The same pixel value would render differently in a different subset. Fine for inspection, but not directly comparable across plots — for that you'd need a fixed value-to-colour mapping.

:::


## Step 8: Zonal statistics → damage zones shapefile

::: {.callout-note}
## QGIS step
The pixel count per zone (`z_count`) was computed in **QGIS** using the Zonal Statistics plugin, applied to `stack_after_before_log.tif` with `beirut_port_explosions_operational_zones_139.shp` as the zone layer. The output — `Raster_cell_count_per_zones.shp` — is the file loaded in the student lab.

This step is done in QGIS because `terra::zonal()` and `terra::extract()` struggled with the file size. If you want to replicate it in R:
:::

```{r}
#| label: zonal-stats-r
#| eval: false  # use QGIS version — kept here for reference only

# R alternative to the QGIS zonal statistics step
# WARNING: very slow on the full-resolution raster

zones_vect <- terra::vect(zones)

zonal_counts <- terra::extract(
  stack_after_before_log,
  zones_vect,
  fun   = "count",
  na.rm = TRUE
)

# Join back to zones and save
zones_with_counts <- zones |>
  mutate(z_count = zonal_counts[, 2])

st_write(zones_with_counts,
         here("data", "geo", "Raster_cell_count_per_zones.shp"),
         delete_dsn = TRUE)
```

## Output files

After running this script the following files should exist:

| File | Used in |
|---|---|
| `data/maxar_processed/stack_after_before_log.tif` | Lab pipeline illustration |
| `assets/stack_after_before_log.png` | Displayed in `lab_part1.qmd` |
| `data/geo/Raster_cell_count_per_zones.shp` | Loaded in `lab_part1.qmd` Section 1.6 |
