---
title: "RF pipeline"
author: "Elisabetta Pietrostefani"
date: "2026-05-19"
---

```{=html}
<div class="part-badge">Lab Appendix</div>
```

This is the **runnable recipe** behind В§1.2 of Lab 1. It walks through the three computational steps that take you from raw MAXAR satellite tiles to a single number per neighbourhood вЂ” the percentage of severely damaged pixels вЂ” using a **supervised Random Forest classifier**.

::: {.callout-important}
## We don't run this in the lab itself

The pipeline is heavy:

-   the raw `.tif` tiles are several GB each
-   training and tuning the Random Forest takes minutes-to-hours depending on the training sample size
-   the full Beirut raster has on the order of **300 million pixels per band**

In the lab we use the **pre-computed output** of this pipeline (`raster_rf_damage.shp`) and focus our time on comparing the satellite-derived damage measure to the survey data (Part 2).

This document is here for the curious вЂ” and for anyone who wants to adapt the workflow to a different event.
:::

The recipe at a glance:

```
2 Г— 3-band rasters   в†’   6-band stacked input   в†’   1 Г— 1-band class raster   в†’   1 polygon per zone
   (pre, post RGB)         (training input)           (severe / medium / none)      (% severe)
        в†‘                         в†‘                          в†‘                           в†‘
   raw MAXAR tiles          Step 1                     Step 2                       Step 3
                       stack + clip                Random Forest                  zonal stats
```

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| eval: false

library(terra)        # raster I/O and operations
library(sf)           # vector I/O
library(dplyr)
library(randomForest) # supervised classifier
library(caret)        # train/test split, confusion matrix
```

------------------------------------------------------------------------

## Step 1 вЂ” Stack the pre- and post-event images

The supervised classifier sees a **single 6-band raster** вЂ” the three pre-event RGB bands and the three post-event RGB bands appended together. Each pixel becomes a point in 6-dimensional spectral space, and the classifier learns boundaries in that space that separate damage classes.

```{r}
#| label: pipeline-step1-stack
#| eval: false

# в”Ђв”Ђ Load the pre-event image в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
# Source: data/maxar_raw/Before/10300500A5F95600.tif (WorldView, 31 Jul 2020)
# 3 bands (R, G, B) В· ~0.5 m pixel size В· many GB

r_before <- terra::rast("data/maxar_raw/Before/10300500A5F95600.tif")

# в”Ђв”Ђ Load the post-event image в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
# Source: data/maxar_raw/After/10300500A6F8AA00.tif (WorldView, 5 Aug 2020)

r_after  <- terra::rast("data/maxar_raw/After/10300500A6F8AA00.tif")

# в”Ђв”Ђ Load the operational zones (used as a mask and later for zonal stats) в”Ђв”Ђв”Ђв”Ђ
# CRS: Deir el Zor (EPSG:22770) вЂ” local Lebanese projection

zones <- sf::st_read(
  "data/geo/beirut_port_explosions_operational_zones_139.shp"
) |>
  sf::st_transform(22770)

zones_vect <- terra::vect(zones)

# в”Ђв”Ђ Align the two acquisitions в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
# The two tiles do not share an identical extent or grid. Resample the "after"
# image onto the "before" grid. method = "near" preserves the original digital
# numbers without interpolating between pixels.

r_after_matched <- terra::resample(r_after, r_before, method = "near")

# в”Ђв”Ђ Crop and mask to the study area в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
# Shrinks file size dramatically and makes the rest of the pipeline tractable.

r_before_crop <- r_before        |> terra::crop(zones_vect) |> terra::mask(zones_vect)
r_after_crop  <- r_after_matched |> terra::crop(zones_vect) |> terra::mask(zones_vect)

# в”Ђв”Ђ Stack: 3 bands (pre) + 3 bands (post) = 6-band input в”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђв”Ђ
# Band order:  [Pre_R, Pre_G, Pre_B, Post_R, Post_G, Post_B]
# This is what the Random Forest will see.

stack_6band <- c(r_before_crop, r_after_crop)
names(stack_6band) <- c("preR", "preG", "preB", "postR", "postG", "postB")

terra::writeRaster(
  stack_6band,
  "data/maxar_processed/stack_6band_pre_post.tif",
  overwrite = TRUE
)
```

::: callout-note
## Why stack rather than difference?

A simpler approach is to compute the log-ratio `log10(post / pre)` per band and feed the classifier the 3-band *change* image. That works, but it throws away information: the classifier never gets to see what the pre-event surface actually looked like, only how much it changed. Stacking the raw bands lets the model learn, for example, that *bright concrete becoming bright rubble* is a different kind of change from *dark asphalt becoming bright rubble* вЂ” both register as "got brighter" in a log-ratio but mean different things on the ground.
:::

------------------------------------------------------------------------

## Step 2 вЂ” Train a supervised Random Forest

This is the heart of the pipeline. We hand the classifier:

1.  the 6-band stack (one row per pixel, six numeric columns)
2.  a set of **training polygons** drawn by an analyst over areas of known damage class

The classifier learns the spectral signature of each class and applies it to every other pixel in the raster.

### 2a. Load training polygons and sample pixels

```{r}
#| label: pipeline-step2a-training
#| eval: false

# Training polygons hand-digitised by the analyst.
# Attribute table contains a `class` column with values:
#   1 = severe damage      (e.g. collapsed warehouses near the port)
#   2 = medium damage      (e.g. blown-out facades, partial roof collapse)
#   3 = little to no damage (untouched residential blocks)

training_poly <- sf::st_read("data/geo/training_polygons.shp") |>
  sf::st_transform(22770)

training_vect <- terra::vect(training_poly)

# Extract every pixel that falls inside a training polygon, together with
# its class label. Each row of `train_pixels` is one pixel:
#   class  preR  preG  preB  postR  postG  postB

train_pixels <- terra::extract(stack_6band, training_vect, bind = TRUE) |>
  as.data.frame() |>
  dplyr::filter(complete.cases(across(preR:postB))) |>
  dplyr::mutate(class = factor(class, levels = c(1, 2, 3),
                               labels = c("severe", "medium", "none")))

# Sanity check: how many training pixels per class?
table(train_pixels$class)
```

### 2b. Train / test split

```{r}
#| label: pipeline-step2b-split
#| eval: false

set.seed(42)

# 70/30 split, stratified by class so each split has the same class proportions

idx <- caret::createDataPartition(train_pixels$class, p = 0.7, list = FALSE)
train_df <- train_pixels[idx, ]
test_df  <- train_pixels[-idx, ]
```

### 2c. Fit the Random Forest

```{r}
#| label: pipeline-step2c-fit
#| eval: false

# Random Forest hyperparameters worth tuning:
#   ntree  вЂ” number of trees. More trees в†’ smaller variance, more compute.
#            Watch the OOB error: stop adding trees when it stops falling.
#   mtry   вЂ” variables sampled at each split. Default sqrt(p) в‰€ 2 for 6 predictors.

rf_model <- randomForest::randomForest(
  class ~ preR + preG + preB + postR + postG + postB,
  data       = train_df,
  ntree      = 500,
  mtry       = 2,
  importance = TRUE
)

# Out-of-bag error: the model's honest internal validation.
# Roughly 1/3 of training pixels are unseen by each tree (the "out-of-bag"
# pixels for that tree). Their predicted labels give an unbiased error estimate
# without needing a separate validation set.
print(rf_model)
plot(rf_model)   # OOB error vs number of trees вЂ” stop where it plateaus

# Variable importance: which bands carry the discriminative signal?
randomForest::varImpPlot(rf_model)
```

::: callout-tip
## When have you trained enough?

Following Rodriguez-Galiano et al. (2012), iteration stops when successive `ntree` increments change the OOB error by less than \~10%. In practice that's usually well below 500 trees for this kind of imagery вЂ” beyond that you're paying compute for negligible accuracy gains.
:::

### 2d. Validate on the held-out test pixels

```{r}
#| label: pipeline-step2d-validate
#| eval: false

test_pred <- predict(rf_model, newdata = test_df)

# Confusion matrix: overall accuracy, per-class precision/recall, kappa
caret::confusionMatrix(test_pred, test_df$class)
```

This is where you find out whether the model generalises. A high overall accuracy that hides a low recall on the "severe" class would be a red flag вЂ” it means the model is good at recognising the *common* class (probably "none") but missing the cases you actually care about.

### 2e. Apply the model to every pixel

```{r}
#| label: pipeline-step2e-predict
#| eval: false

# `terra::predict()` runs the model in chunks so the full raster doesn't have
# to be held in RAM at once. Output is a single-band raster where each cell
# holds the predicted class (1 / 2 / 3).

damage_class <- terra::predict(
  stack_6band,
  model    = rf_model,
  na.rm    = TRUE,
  filename = "data/maxar_processed/raster_rf_damage.tif",
  overwrite = TRUE
)

# Quick visual sanity check
plot(damage_class, col = c("red", "orange", "lightyellow"),
     main = "Predicted damage class (1 = severe, 2 = medium, 3 = none)")
```

------------------------------------------------------------------------

## Step 3 вЂ” Aggregate pixels to operational zones

We have a damage class for every \~0.5 m pixel. For statistical analysis and joining to other data sources (survey, census, deprivation index) we need a number *per zone*, not per pixel.

```{r}
#| label: pipeline-step3-zonal
#| eval: false

# Total pixels per zone (denominator)
n_total  <- terra::extract(
  damage_class, zones_vect,
  fun = function(x) sum(!is.na(x))
)

# Severe-damage pixels per zone вЂ” class id 1
n_severe <- terra::extract(
  damage_class, zones_vect,
  fun = function(x) sum(x == 1, na.rm = TRUE)
)

zones$n_severe_px <- n_severe[, 2]
zones$n_total_px  <- n_total[, 2]
zones$pct_severe  <- 100 * zones$n_severe_px / zones$n_total_px

# This is the file we load in В§1.6 of the lab
sf::st_write(zones, "data/geo/raster_rf_damage.shp", delete_dsn = TRUE)
```

::: callout-tip
## Sense-check the orders of magnitude

A zone of 0.2 kmВІ at 0.5 m resolution contains roughly **800,000 pixels**. If `n_severe_px` for that zone comes back as \~96,000, then `pct_severe в‰€ 12 %`. If it comes back as 0 or as 800,000, something has gone wrong вЂ” either a class id is mismatched, or the mask did not align with the zone boundary, or the zone is mostly sea. Always sanity-check denominators on a known zone first.
:::

------------------------------------------------------------------------

## Recap: what's actually on disk after each step

| Step | File written | Shape / dimensionality |
|---------------|----------------------------------|------------------------|
| 1    | `stack_6band_pre_post.tif`       | 6 bands В· \~0.5 m pixels       |
| 2    | `raster_rf_damage.tif`           | 1 band В· \~0.5 m pixels В· values в€€ {1, 2, 3} |
| 3    | `raster_rf_damage.shp`           | 139 polygons В· `pct_severe` attribute |

The last file is the one that goes into Lab 1, В§1.6 вЂ” and it's the variable we compare against the survey in Part 2.

------------------------------------------------------------------------

## Limitations to keep in mind

The same five limitations from Lab 1 В§1.2 apply to anything derived from this pipeline:

-   **Line-of-sight bias** вЂ” at nadir the satellite sees rooftops, not facades.
-   **Rubble removal** вЂ” pixels cleared before the post-image read as undamaged.
-   **Cloud and smoke** вЂ” optical sensors can't see through either.
-   **Temporal gap** вЂ” the post-image is \~24 h after the blast; some pixels have already shifted.
-   **The 0.5 m floor** вЂ” anything smaller than the pixel is invisible.

And one specific to the supervised approach used here:

-   **Training-label bias** вЂ” the classifier only knows the damage signatures it was shown. If training polygons over-represent collapsed warehouses near the port and under-represent partial-facade damage further out, the classifier will systematically under-call damage in residential areas. The OOB error tells you how well the model predicts *its own training distribution*, not how well it generalises.

This last point is the one most likely to bite in the Part 2 comparison with the survey. If the survey reports more damage than the satellite in residential blocks far from the port, training-label bias is one of the prime suspects.