---
title: "Part 3 · Damage, sentiment & economic stress"
author: "Elisabetta Pietrostefani, University of Liverpool"
date: "2026-05-19"
---

```{=html}
<div class="part-badge">Lab Part 3 of 3</div>
```

::: objectives
**Learning objectives**

By the end of Part 3 you will be able to:

-   Run a simple OLS model linking building-level damage to a household-level outcome
-   Interpret the slope on a damage variable as a "one-grade-worse" effect
-   Identify which alternative damage measures and outcomes are worth exploring next
:::

::: callout-important
## Quick reminder: the survey is synthetic

The blast damage variables (`windows_rev`, `walls_rev`, `furniture_rev`, `BLASTav`) are preserved as collected. Everything else — including the three outcomes we use below — has been re-sampled. The marginal distributions look right and the link to building damage is preserved, but **do not use these results for inference beyond the lab**.
:::

------------------------------------------------------------------------

## 3.1 Why this part is light

Parts 1 and 2 spent a long time building, comparing and validating damage measures. Part 3 takes the simplest possible next step: does damage actually *predict anything we care about* at the household level?

We'll run **one regression** linking the Copernicus structural grade (`dmg_grd`) to three household outcomes, then hand the rest of the analysis over to you.

------------------------------------------------------------------------

## 3.2 Setup and join

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(tidyverse)
library(readxl)
library(broom)       # tidy regression output
library(knitr)
```

We need the survey at household level with each respondent's *building damage grade* attached. The 2021 wave only, across all three neighbourhoods (Ras Beirut included).

```{r}
#| label: join-survey-buildings
#| eval: false

samplings <- readxl::read_excel("data/kr_rb_mm_synthetic.xlsx") |>
  dplyr::mutate(
    windows_rev   = 5 - windows_rev,
    walls_rev     = 5 - walls_rev,
    furniture_rev = 5 - furniture_rev,
    BLASTav       = 5 - BLASTav
  )

buildings <- sf::st_read("data/geo/buildings_damaged.shp", quiet = TRUE) |>
  sf::st_drop_geometry() |>
  dplyr::select(bld_nbr, dmg_grd)

hh <- samplings |>
  dplyr::filter(year == 2021) |>
  dplyr::left_join(buildings, by = "bld_nbr") |>
  # ordered factor: severity rises from "no damage" to "destroyed"
  dplyr::mutate(dmg_grd = factor(
    dmg_grd,
    levels  = c("no damage", "minor damage", "moderate damage",
                "severe damage", "destroyed"),
    ordered = TRUE
  )) |>
  dplyr::filter(!is.na(dmg_grd))

nrow(hh)
```

------------------------------------------------------------------------

## 3.3 Three outcomes, one quick look

The three synthetic outcomes we'll use:

| Variable | Description | Scale |
|---|---|---|
| `hh_income_dollars` | Self-reported monthly household income | USD |
| `nation_hh` | Household nationality | 1 = Lebanese, 2 = Syrian, 3 = Other, 100 = didn't reply |
| `confid_municipality` | Confidence in the local municipality | 1 = none → 5 = a lot |

`nation_hh` is **categorical**, not ordinal — averaging it gives a meaningless number. For the descriptive look below we recode it as a binary (% Lebanese) and treat 100s as NA. We'll come back to it in §3.5 as a possible *control* or *split* variable rather than an outcome.

```{r}
#| label: descriptives-by-damage
#| eval: false

hh <- hh |>
  dplyr::mutate(
    nation_hh = dplyr::na_if(nation_hh, 100)   # 100 = didn't reply → NA
  )

hh |>
  dplyr::group_by(dmg_grd) |>
  dplyr::summarise(
    n                = dplyr::n(),
    mean_income      = mean(hh_income_dollars,   na.rm = TRUE),
    pct_lebanese     = 100 * mean(nation_hh == 1, na.rm = TRUE),
    mean_confid_muni = mean(confid_municipality, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(
    digits  = 2,
    caption = "Three outcomes by Copernicus damage grade"
  )
```
------------------------------------------------------------------------

## 3.4 One regression

We treat the ordered damage grade as a numeric severity score (0 = no damage → 4 = destroyed) and regress monthly income on it. A simple, single specification — no controls, no fixed effects, no heteroskedasticity-robust SEs. Just the headline relationship.

```{r}
#| label: damage-income-regression
#| eval: false

hh <- hh |> dplyr::mutate(dmg_grd_num = as.integer(dmg_grd) - 1L)

m_income <- lm(hh_income_dollars ~ dmg_grd_num, data = hh)

broom::tidy(m_income) |>
  knitr::kable(digits = 2,
               caption = "Effect of a one-grade increase in damage on monthly income")
```

::: callout-note
## Reading the regression

- **Slope = −75 USD, p < 0.001.** Each one-grade-worse rating on the Copernicus damage scale predicts about $75 less in self-reported monthly income. A household in a *severely damaged* building (grade 3) sits roughly $225 below a household in an *undamaged* building (grade 0).

- **Intercept ≈ $264.** This is the model's predicted monthly income for an undamaged-building household — a sobering anchor that reflects Beirut's economic collapse rather than anything about the blast itself. Most respondents are reporting incomes that put them well below any reasonable poverty line.

- **n ≈ 2,790.** Much larger than the zone-level regression in Part 2 (n = 18), so the standard error is tight and the slope is robust.

**This is correlation, not causation.** Three stories are consistent with this slope:

1. The blast destroyed earning capacity for households in heavily damaged buildings (causal, in the direction you'd hope to claim).
2. Poorer households were already living in the most vulnerable buildings before the blast (selection — the buildings most likely to be damaged were the ones already housing the poorest people).
3. Some unobserved third variable (neighbourhood, housing tenure, age structure) drives both poverty *and* building vulnerability.

You can't separate these with a single cross-sectional regression. The §3.5 exploration prompts give you ways to start chipping at them.
:::

------------------------------------------------------------------------

## 3.5 Your turn — explore

You now have the machinery to ask a lot of questions in a few lines of code. Some directions to try:

**Swap the damage variable.** `dmg_grd_num` is one of at least four damage measures you have in hand. The full set:

- `dmg_grd_num` — Copernicus expert grade (what we just used)
- `dmg_JRC_num` — JRC automated grade (much more conservative)
- `BLASTav` — household self-report (interior damage)
- `pct_severe_adj` — satellite RF % severe at zone level (need to join via §2.6 lookup)

Rerun the regression with each. Does the coefficient stay the same sign? Same magnitude? Where do they diverge most, and why?

**Swap the outcome.** Try `nation_hh` or `confidence_municipality` instead of income. Does damage predict sentiment as well as it predicts earnings? Does it predict one but not the other? *Hypothesis to test:* damage to your home might shake confidence in local institutions (who's supposed to rebuild?) more than it shakes feelings about the nation.

**Add a control or many.** Neighbourhood is the obvious one:

```r
lm(hh_income_dollars ~ dmg_grd_num + neighbourhood, data = hh)
```

This asks: *does damage still predict income once we account for the fact that Karantina is generally poorer than Ras Beirut?* If the damage coefficient shrinks dramatically, damage was mostly proxying for which neighbourhood the household was in. If it stays, damage carries signal *within* neighbourhoods.

**Compare survey vs structural damage as predictors.** Run two regressions side by side: one with `dmg_grd_num` (what an expert saw from imagery) and one with `BLASTav` (what the household reported about their own home). Which one predicts income better? Both points back to a deeper question: *which damage measure is a better proxy for the actual harm done to a household?*

------------------------------------------------------------------------

::: callout-tip
## Things worth discussing

- The synthetic data preserves the *correlation* between damage and outcomes but breaks the joint structure between non-damage variables. What kinds of questions can we still ask honestly? What kinds become unreliable?
- We don't have a pre-blast measure of income in 2021 (only 2018, with a different sample). What kind of identification strategy would let you say something causal about the blast's effect on income? What data would you need?
- If satellite damage predicts income with roughly the same coefficient as Copernicus damage, that's good news for humanitarian work: you could target relief from satellite imagery alone, without sending surveyors into the field. Under what conditions would that conclusion be wrong?
:::
